---
title: Assignment - Sustainable Hunting Model
toc: true
format:
  html:
    embed-resources: true
    code-fold: false 
    code-tools: true 
    mathjax: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: sentence
author: 
  - name: Karol Paya & Oksana Protsukha
    affiliation: Master of Environmental Data Science Program
date: 2025-06-01
---

## PART 1: Test the model with best guess parameters

Build this model (e.g add hunting to the lotvmodK.R),

Some requirements/hints for your model

-   You should make sure that you don't hunt more prey than exist.

-   To ensure that you might also add a minimum prey population input that must be met before hunting is allowed.

-   Note you can make this as simple or as complex as you would like. You could represent hunting in a way that is similar to "harvesting" in the last assignment. 

```{r, setup,  message=FALSE, warning=FALSE, results='hide'}
#| eval: true
#| echo: false
#| code-fold: true


#Load libraries
library(deSolve)
library(tidyverse)
library(ggplot2)
library(tidyr)
library(dplyr)
library(purrr)
library(sensitivity)
library(ggpubr) # to arrange plots

# Source function
source("lotvmodK_hunt_v2.R")
```


```{r}
# Initial conditions
currpop <- c(prey = 80, pred = 15)

# Time points to see results
times <- seq(0, 100, by = 1)

# Model parameters
pars <- c(
  rprey = 0.95, 
  K = 2000, 
  alpha = 0.01, 
  pmort = 0.4,
  eff = 0.6,
  hunt_rate = 0, 
  min_prey = 60
)

# Run the model
res <- ode(func = lotvmodK_hunt_v2, y = currpop, times = times, parms = pars)

# Visualize results
resl <- as.data.frame(res) %>% 
  pivot_longer(-time, names_to = "pop_type", values_to = "pop")

p1 <- ggplot(resl, aes(time, pop, col = pop_type)) +
  geom_line() +
  labs(title = "Prey-Preditor Interaction with Hunting")

p2b <- ggplot(as.data.frame(res), aes(pred, prey, col = time)) +
  geom_point() +
  labs(y = "Prey", x = "Predators")

ggarrange(p1, p2b)

```


## PART 2

# Part 2.1: Explore System Stability Using Scenario-Based Approach

### Define Stability
**Stability** is defined at the intersection of dprey/dt = 0 and dpred/dt = 0.
Sustainable hunting maintains both populations above minimum thresholds.

```{r, scenario_based_stability, message=FALSE, warning=FALSE, results='hide'}
#| eval: true
#| echo: true
#| code-fold: true


# Base parameters
base_pars <- list(
  rprey = 0.95,
  alpha = 0.01,
  eff = 0.6,
  pmort = 0.4,
  K = 2000
)

# Define hunting scenarios
scenarios <- data.frame(
  scenario = c("No Hunting (0%)", "Low Hunting (5%)", "Medium Hunting (10%)", "High Hunting (20%)"),
  hunt_rate = c(0, 0.05, 0.10, 0.20),
  min_prey = c(0, 50, 400, 500)
)

print(scenarios)

# Function to find equilibrium for one scenario
find_single_equilibrium <- function(hunt_rate, min_prey) {
  
  # Search ranges
  prey_vals <- seq(200, 1200, by = 100)
  pred_vals <- seq(50, 200, by = 25)
  
  best_distance <- Inf
  best_prey <- NA
  best_pred <- NA
  
  pars <- c(base_pars, hunt_rate = hunt_rate, min_prey = min_prey)
  
  for (prey in prey_vals) {
    for (pred in pred_vals) {
      pop <- c(prey = prey, pred = pred)
      derivs <- lotvmodK_hunt_v2(t = 0, pop = pop, pars = pars)
      
      dprey <- derivs[[1]][1]
      dpred <- derivs[[1]][2]
      
      distance <- sqrt(dprey^2 + dpred^2)
      
      if (distance < best_distance) {
        best_distance <- distance
        best_prey <- prey
        best_pred <- pred
      }
    }
  }
  
  return(list(prey = best_prey, pred = best_pred, distance = best_distance))
}

# Find equilibria for all scenarios
results <- data.frame()

for (i in 1:nrow(scenarios)) {
  cat("Finding equilibrium for", scenarios$scenario[i], "...\n")
  
  eq <- find_single_equilibrium(scenarios$hunt_rate[i], scenarios$min_prey[i])
  
  # Calculate harvest
  harvestable <- max(0, eq$prey - scenarios$min_prey[i])
  annual_harvest <- scenarios$hunt_rate[i] * harvestable
  
  # Check sustainability
  sustainable <- (eq$prey > 100) & (eq$pred > 10)
  
  result_row <- data.frame(
    scenario = scenarios$scenario[i],
    hunt_rate = scenarios$hunt_rate[i],
    min_prey = scenarios$min_prey[i],
    equilibrium_prey = eq$prey,
    equilibrium_pred = eq$pred,
    annual_harvest = round(annual_harvest, 0),
    sustainable = sustainable
  )
  
  results <- rbind(results, result_row)
}

print(results)

# Find best sustainable option
sustainable_options <- results[results$sustainable == TRUE, ]

if (nrow(sustainable_options) > 0) {
  best <- sustainable_options[which.max(sustainable_options$annual_harvest), ]
  
  print("\n Optimal Parameters for Sustainable Hunting")
  print(paste("Scenario:", best$scenario))
  print(paste("Hunt Rate:", best$hunt_rate * 100, "%"))
  print(paste("Minimum Prey Threshold:", best$min_prey))
  print(paste("Equilibrium Prey:", best$equilibrium_prey))
  print(paste("Equilibrium Predators:", best$equilibrium_pred))
  
} else {
  print("All scenarios lead to unsustainable population levels.")
}

# Simple stability plot (like your forest example)
prey_range <- seq(100, 1000, by = 50)
stability_data <- data.frame()

for (i in 1:nrow(scenarios)) {
  fixed_pred <- results$equilibrium_pred[i]  # Use equilibrium predator level
  pars <- c(base_pars, hunt_rate = scenarios$hunt_rate[i], min_prey = scenarios$min_prey[i])
  
  for (prey in prey_range) {
    pop <- c(prey = prey, pred = fixed_pred)
    derivs <- lotvmodK_hunt_v2(t = 0, pop = pop, pars = pars)
    dprey <- derivs[[1]][1]
    
    stability_data <- rbind(stability_data, data.frame(
      scenario = scenarios$scenario[i],
      prey = prey,
      dprey = dprey
    ))
  }
}

# Plot stability analysis
p <- ggplot(stability_data, aes(x = prey, y = dprey, color = scenario)) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Prey Population Stability Analysis",
       subtitle = "Equilibrium where dprey/dt = 0 (dashed line)",
       x = "Prey Population", 
       y = "Rate of Change (dprey/dt)",
       color = "Hunting Scenario") +
  theme_minimal()

print(p)

```

# Part 2.2: Find Sustainable Hunting Rate Using Optimization Approach

### Define Stability

**Stability** is defined as mean relative deviation from natural equilibrium over the final 3 years, assuming populations remain within viable recovery thresholds.

### Calculate Optimization Parameters for Sustainable Hunting

```{r, stability_function, message=FALSE, warning=FALSE, results='hide'}
#| eval: true
#| echo: true
#| code-fold: true

# calculate stability for given parameters
calculate_stability <- function(params, natural_prey, natural_pred) {
  # Run model with given parameters
  result <- ode(y = c(prey = 500, pred = 50),
               times = seq(1, 365 * 8, by = 5),
               func = lotvmodK_hunt_v2,
               parms = params)
  
  # Get last 3 years and calculate mean deviation from equilibrium
  last_3_years <- tail(as.data.frame(result), 219)
  
  prey_deviation <- mean(abs(last_3_years$prey - natural_prey)) / natural_prey
  pred_deviation <- mean(abs(last_3_years$pred - natural_pred)) / natural_pred
  
  return((prey_deviation + pred_deviation) / 2)
}

# Base ecological parameters
base_pars <- list(rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.4, K = 2000)

# Find natural equilibrium (no hunting)
cat("Calculating natural equilibrium...\n")
natural_result <- ode(y = c(prey = 500, pred = 50), 
                     times = seq(1, 365 * 10, by = 5),
                     func = lotvmodK_hunt_v2, 
                     parms = c(base_pars, hunt_rate = 0, min_prey = 0))

equilibrium_data <- tail(as.data.frame(natural_result), 146)
natural_prey <- mean(equilibrium_data$prey)
natural_pred <- mean(equilibrium_data$pred)

# Generate scenarios
set.seed(123)
n <- 1800

scenarios <- data.frame(
  rprey = runif(n, 0.9, 1.0),
  alpha = runif(n, 0.009, 0.011),
  eff = runif(n, 0.55, 0.65),
  pmort = runif(n, 0.38, 0.42),
  K = runif(n, 1900, 2100),
  hunt_rate = runif(n, 0, 0.3),
  min_prey = runif(n, 40, 80)
)

# Calculate stability for each scenario
stability_scores <- numeric(nrow(scenarios))
for (i in 1:nrow(scenarios)) {
  stability_scores[i] <- calculate_stability(as.list(scenarios[i, ]), natural_prey, natural_pred)
}

# Find best scenario
best_idx <- which.min(stability_scores)
best_params <- scenarios[best_idx, ]

# Get final populations for best scenario
final_result <- ode(y = c(prey = 500, pred = 50),
                   times = seq(1, 365 * 8, by = 5),
                   func = lotvmodK_hunt_v2,
                   parms = as.list(best_params))
final_pops <- tail(as.data.frame(final_result), 1)

```


```{r, resuts, message=FALSE, warning=FALSE,}
#| eval: true
#| echo: false
#| code-fold: false

# Display results
cat("\nNATURAL EQUILIBRIUM (no hunting):\n")
cat("Prey:", round(natural_prey, 1), "\n")
cat("Predators:", round(natural_pred, 1), "\n")

cat("\nOPTIMAL HUNTING STRATEGY:\n")
cat("Hunt Rate:", round(best_params$hunt_rate * 100, 1), "%\n")
cat("Min Prey Threshold:", round(best_params$min_prey, 1), "\n")
cat("Final Prey:", round(final_pops$prey, 1), "\n")
cat("Final Predators:", round(final_pops$pred, 1), "\n")
cat("Stability Score:", round(stability_scores[best_idx], 4), "(lower = more stable)\n")

cat("\nStability = mean deviation from natural equilibrium over last 3 years assuming prey and pred populations remain within recovery thresholds.\n")

```

### Visualize Results

```{r, sobol_viz, stability_function, message=FALSE, warning=FALSE}
#| eval: true
#| echo: true
#| code-fold: true

times <- seq(1, 365, by = 1)
currpop <- c(prey = 500, pred = 50)

# Define 4 hunting scenarios
scenarios <- data.frame(
  scenario = c("No Hunting", "Low Hunting", "Medium Hunting", "High Hunting"),
  hunt_rate = c(0, 0.1, 0.2, 0.3),
  min_prey = c(0, 45, 50, 55)
)

# Create time series plots for all scenarios
plot_list <- list()

for (i in 1:nrow(scenarios)) {
  pars <- c(base_pars, hunt_rate = scenarios$hunt_rate[i], min_prey = scenarios$min_prey[i])
  currpop <- currpop
  times <- times
  
  # Run the model
  res <- ode(func = lotvmodK_hunt_v2, y = currpop, times = times, parms = pars)
  
  # Visualize results
  resl <- as.data.frame(res) %>% pivot_longer(-time, names_to = "pop_type", values_to = "pop")
  
  p <- ggplot(resl, aes(time/365, pop, col = pop_type)) +
    geom_line(size = 1.2) +
    scale_color_manual(values = c("prey" = "lightgreen", "pred" = "red")) +
    labs(title = paste(scenarios$scenario[i], "- Hunt Rate:", scenarios$hunt_rate[i]*100, "%"),
         x = "Years", y = "Population", color = "Species") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  plot_list[[i]] <- p
}

# PLot all time series plots in a 2x2 grid
combined_grid <- ggarrange(plotlist = plot_list, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")
print(combined_grid)

# Summary table
summary_data <- data.frame()
for (i in 1:nrow(scenarios)) {
  pars <- c(base_pars, hunt_rate = scenarios$hunt_rate[i], min_prey = scenarios$min_prey[i])
  currpop <- currpop
  times <- times
  
  res <- ode(func = lotvmodK_hunt_v2, y = currpop, times = times, parms = pars)
  
  # Get final 2 years average
  final_data <- tail(as.data.frame(res), 146)  # ~2 years of data
  
  summary_data <- rbind(summary_data, data.frame(
    scenario = scenarios$scenario[i],
    hunt_rate = scenarios$hunt_rate[i] * 100,
    avg_prey = round(mean(final_data$prey), 1),
    avg_pred = round(mean(final_data$pred), 1)
  ))
}

print(summary_data)
```

